On the Gravitational stability of gravito-turbulent accretion disks

### 2016-01-25 paper/summary.tex
rough version 2, old figs

However, \cite{baehr15} included a floor temperature (or irradiation)
in their simulations but still find that at fixed cooling rates, disks
eventually fragment with sufficient spatial resolution.
In light of our results on viscous GI, 
we here highlight a possible contribution: %in addition to num visc (meru lodato)
high resolutions enable small-scale turbulent angular momentum transport 
% physical, cf numerical  
to aid clump formation (see also \S\ref{MHD} below). 
This mechanism is unavailable at low resolutions, for which one must
rely on rapid cooling.

@@

We note that, although power in gravito-turbulence is dominated by scales with
$kH\sim 1$ \citep{cossins09}, which is  
well-resolved by modern numerical simulations, that this does not
necessarily imply small-scale dynamics/thermodynamics are unimportant
\citep[especially for non-linear evolution, see][]{young15},  
as is evident from the non-converging simulations described above.

@@

\subsection{MHD turbulence}\label{MHD}%turbulence and viscosity 
We emphasize the linear theory we have developed does not assume

%It is important to note that while we consider gravito-turbulent disks  
%as an application, the linear theory itself does not assume any 
%specific origin of turbulence.

%effect of MHD turbulence on GI 
Our models may thus apply to self-gravitating disks dominated by, for
example, magneto-hydrodynamic (MHD) turbulence.   
%In fact, the viscous approach may be more
%applicable here than gravito-turbulence, because the characteristic
%wavelength of the magneto-rotational instability (which leads to MHD
%turbulence) is $\ll H$, at least in the ideal MHD limit. Then there is
%a scale separation between the underlying, small-scale MHD 
%turbulence and the large-scale effects of self-gravity. 
%fromang is isothermal sims - no cooling effect 
Explicit numerical simulations of magnetized, massive disks have been performed 
by \cite{fromang05}. He finds disk fragmentation with increasing numerical
resolution, attributing this to resolving the most unstable MRI
wavelength, which enables small-scale angular momentum removal by MHD
turbulence to aid fragmentation. %this supports using viscosity to
                                %mimic turbulence
This would be a physical mechnism represented by the viscous GI
discussed in this paper; which lends some support for  
the use of a viscosity to represent turbulence.

@@

\subsection{Outstanding issues}
%the link between the linear GI's discussed in this paper 
%and frag 
True disk fragmentation is a non-linear process characterized by clumps   
reaching densities orders of magnitude above the ambient value, and
which survives tidal shear and shocks \citep{paardekooper12,young16}.  
%cite the young papers 
Clearly, our linear models cannot address fragmentation
directly. Technically, we have only demonstrated that a 
gravito-turbulent state becomes dynamically unstable, and thus should
not persist, when cooling (viscosity) is too rapid (large).  
However, numerical simulations show that there are only two
possible outcomes of cooling, self-gravitating disks: steady
gravito-turbulence or fragmentation. It seems then reasonable to 
speculate that the non-existence of a gravito-turbulent state, here due to
dynamical instability, implies disk fragmentation.

{\bf comment on stochasitc frag. we only consider 
how to enhance GI by reducing rotation and pressure. but a third route is 
increasing sigma. according to toomre Q these are the 3 effects one should consider
for SG. turbulent flucations may occasionally produce high sigma dens perts. 
possible relation to current framework? high sigma perts arise because of viscous GI? 
high alpha $\to$ small-scale transport $\to$ enhance over-dense regions}

@@

%non-axisymmetry and transient dynamics 
Another possible generalization is non-axisymmetric
disturbances. In barotropic, inviscid disks  
non-axisymmetric GI can develop for $Q$ somewhat larger than unity
\citep{papaloizou89,papaloizou91}. It would be interesting to study
how non-axisymmetric perturbations are affected by cooling and
viscosity in order to improve %or verify 
the link between disk fragmentation and the
stability of gravito-turbulent disks.

>>>>

On the other hand, \cite{baehr15} included a floor temperature (or irradiation)
in their simulations but still find that at fixed cooling rates, disks
eventually fragment with sufficient spatial resolution.
In light of our results on viscous GI, 
we here highlight a possible contribution: %in addition to num visc (meru lodato)
high resolutions enable small-scale turbulent angular momentum transport 
% physical, cf numerical  
to aid clump formation (see also \S\ref{MHD} below). 
This mechanism is unavailable at low resolutions, for which one must
rely on rapid cooling.

@@

We comment that although modern simulations resolve the dominant
scale associated with gravito-turbulence, $kH\sim 1$
\citep{cossins09}, this does not
necessarily imply small-scale dynamics/thermodynamics are unimportant 
\citep[especially for non-linear evolution, see][]{young15},  
as is evident from the non-converging simulations described above.

@@

\subsection{MHD turbulence}\label{MHD}%turbulence and viscosity 
We emphasize the linear theory we have developed does not assume  
a particular origin for the turbulence represented by the imposed
viscosity. For example, the 2D dispersion relation, Eq. \ref{thindisk}
and Appendix \ref{2ddisp}, treats $\alpha$ as an independent
input parameter.

%We only considered gravito-turbulence as an application.  
%In
%practice, this amounts to relating the $\alpha$ parameter to $Q$. 
%most equations treat alpha/Q/beta as independent

%effect of MHD turbulence on GI 
Our models may thus apply to self-gravitating disks dominated by MHD
turbulence.     
%In fact, the viscous approach may be more
%applicable here than gravito-turbulence, because the characteristic
%wavelength of the magneto-rotational instability (which leads to MHD
%turbulence) is $\ll H$, at least in the ideal MHD limit. Then there is
%a scale separation between the underlying, small-scale MHD 
%turbulence and the large-scale effects of self-gravity. 
%fromang is isothermal sims - no cooling effect 
Explicit numerical simulations of magnetized, massive disks have been performed 
by \cite{fromang05}. He finds disk fragmentation with increasing numerical
resolution, attributing this to resolving the most unstable MRI
wavelength, which enables small-scale angular momentum removal by MHD
turbulence to aid fragmentation. %this supports using viscosity to
                                %mimic turbulence
This would be a physical mechnism represented by the viscous GI
discussed in this paper; which lends some support for  
the use of a viscosity to represent turbulence.

@@

\subsection{Outstanding issues}
%the link between the linear GI's discussed in this paper 
%and frag 
True disk fragmentation is a non-linear process characterized by
clumps reaching densities orders of magnitude above the ambient value, and
which survives tidal shear and shocks \citep{paardekooper12,young16}.  
%cite the young papers 
Clearly, our linear models cannot address fragmentation
directly. Technically, we have only demonstrated that a 
gravito-turbulent state becomes dynamically unstable, and thus should
not persist, when cooling is too rapid or when the associated viscous
stresses are too large.  However, numerical simulations show that
there are only two possible outcomes of cooling, self-gravitating
disks: steady gravito-turbulence or fragmentation. It seems then
reasonable to  speculate that the non-existence of a stable
gravito-turbulent state, here due to dynamical instability, implies
disk fragmentation.

%Our goal is to highlight how non-ideal effects, such as viscosity and
%cooling, can promote GI. However, the Toomre parameter
%(Eq. \ref{toomreQ_criterion}) suggest a third possibility: increasing
%the surface density.

Our deterministic approach cannot model `stochastic fragmentation'
\citep{paardekooper12,hopkins13}. In this interpretation,  
fragmentation is attributed to the occurence (and survival) 
of non-linear density perturbations, which arise from the
gravito-turbulent state simply by chance. 
%Whether these high
%density perturbations actually lead to a clump formation, 
This might be considered as a secondary instability by lowering the
Toomre $Q$ through a (random) increase in density
(Eq. \ref{toomreQ_criterion}). By contrast, the viscosity/cooling
mechanisms discussed in this paper lowers $Q$ by reducing pressure and
Coriolis forces. %we have a physical way of lowering pressure/rotation

%stochasitc frag -> it just happens by chance. no physics

@@

%non-axisymmetry and transient dynamics 
%While our linear problem has been formulated with some 
Another possible generalization is non-axisymmetric
disturbances. In barotropic, inviscid disks  
non-axisymmetric global GI can develop for $Q$ somewhat larger than unity
\citep{papaloizou89,adams89,papaloizou91}. It would be interesting to study
how non-axisymmetric perturbations are affected by cooling and
viscosity in order to improve %or verify 
the link between disk fragmentation and the
stability of gravito-turbulent disks.
